---
title: "R Notebook"
output: html_notebook
---
```{r}
setwd("/Users/emmasainovic")
finance_data <- read.csv("/Users/emmasainovic/Downloads/dida 130 (winter 2025)/final/finance.csv")
finance <- finance_data[complete.cases(finance_data),]


library(dplyr)
library(car)
library(ggplot2)
```

Research Question : In New York State, how do ‘Details.Health.Health Total Expenditure’, ‘Details.Education.Education Total’, ‘Details.Natural Resources.Parks.Parks Total Expenditure’ and ‘Details.Correction.Correction Total’ impact the ‘Totals.Expenditure’ between 1992 and 2019? 
```{r}
ny <- finance %>% filter(State == 'NEW YORK')
ny

# regression model 
ny_reg <- lm(Totals.Expenditure
 ~ Details.Health.Health.Total.Expenditure
 + Details.Education.Education.Total
 + Details.Natural.Resources.Parks.Parks.Total.Expenditure
 + Details.Correction.Correction.Total,
 data = ny
)
ny_sum <- summary(ny_reg)
ny_sum

# checking assumptions of regression model

# normality of residuals
shapiro.test(residuals(ny_reg))

# homogenous
ncvTest(ny_reg)

# colinearity 
vif(ny_reg)

# linearity
plot(ny_reg)
residualPlots(ny_reg)

# statistics summary table
mean <- ny%>%summarize(across(c(Totals.Expenditure,Details.Health.Health.Total.Expenditure,Details.Education.Education.Total,Details.Natural.Resources.Parks.Parks.Total.Expenditure,Details.Correction.Correction.Total), ~mean(., na.rm=T)))

min <- ny%>% summarise(across(c(Totals.Expenditure,Details.Health.Health.Total.Expenditure,Details.Education.Education.Total,Details.Natural.Resources.Parks.Parks.Total.Expenditure,Details.Correction.Correction.Total), ~min(., na.rm=T)))

max <- ny%>% summarise(across(c(Totals.Expenditure,Details.Health.Health.Total.Expenditure,Details.Education.Education.Total,Details.Natural.Resources.Parks.Parks.Total.Expenditure,Details.Correction.Correction.Total), ~max(., na.rm=T)))

sd <- ny%>% summarise(across(c(Totals.Expenditure,Details.Health.Health.Total.Expenditure,Details.Education.Education.Total,Details.Natural.Resources.Parks.Parks.Total.Expenditure,Details.Correction.Correction.Total), ~sd(., na.rm=T)))

iqr <- ny%>% summarise(across(c(Totals.Expenditure,Details.Health.Health.Total.Expenditure,Details.Education.Education.Total,Details.Natural.Resources.Parks.Parks.Total.Expenditure,Details.Correction.Correction.Total), ~IQR(., na.rm=T)))

sum_table <- rbind(mean, min, max, sd, iqr)
rownames(sum_table) <- c('mean',' minimum', 'maximum', 'standard deviation', 'IQR')
sum_table

# data frame for bar graph of mean values of variables 
mean_health <- mean(ny$Details.Health.Health.Total.Expenditure)
mean_edu <- mean(ny$Details.Education.Education.Total)
mean_nat <- mean(ny$Details.Natural.Resources.Parks.Parks.Total.Expenditure)
mean_cor <- mean(ny$Details.Correction.Correction.Total)

mean_vals <- c(mean_health, mean_edu, mean_nat, mean_cor)
mean_labs <- c('health', 'education','recreational','correctional')

means_data <- data.frame(mean_labs, means_val)

# bar graph 
ggplot(means_data, aes(y = mean_vals, x = mean_labs, fill=mean_labs))+
  geom_bar(stat = "identity") +
  labs(x = "Category of Taxes dollars spent", y = "Amount Spent",
       title = "Tax dollars spent in NY state between 1992 and 2019 by category")+
  theme(legend.position = "none")

# histogram 1
hist_ny1 <-ggplot(ny, aes(Details.Education.Education.Total))+
  geom_histogram(fill = "pink", color = "black")+
  labs(x = "Spending on Education", y = "Count of Years")+
  theme_minimal()
hist_ny1

# histogram 2
hist_ny2 <-ggplot(ny, aes(Totals.Expenditure))+
  geom_histogram(fill = "lightblue", color = "black")+
  labs(x = "Spending on Nature", y = "Count of Years")+
  theme_minimal()
hist_ny2

# line graph for total spent
total_spent <- ggplot(data = ny, aes(y = Totals.Expenditure, x = Year))+
  labs(x = 'Year' , y = "$ Spent", title = "New York State Total Expenditure, 1992-2019 ")+
geom_line()
total_spent

# line graph for money spent for education
edu_spent <- ggplot(data = ny, aes(y = Details.Education.Education.Total, x = Year))+
  labs(x = 'Year' , y = "$ Spent", title = "New York State Educational Expenditure, 1992-2019 ")+
  geom_line()
edu_spent

```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

